Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2018-10-28 mprop/mprop.tex
Just before I rewrite everything

  \mathcal{L}(\bm\nu) &= \frac{1}{2}\log|\bm\Sigma| + \frac{1}{2}\log|\mathbf{S}| + \mathbb{E}\left[-\frac{1}{2}\Tr[\Kuu^{-2}] -
    \sum_{i=1}^d \log(\lambda_i + 1)\right] \\

@@

\cite{DBLP:books/lib/Bishop07}. This allows us to
simplify the expression to the following:

@@

  &+ \sum_{i=1}^N \sum_{t=1}^T \mathbb{E}[r(s_{i,t})] - \mathbb{E}[V_r(s_{i,t})] + \gamma\sum_{s' \in \mathcal{S}} \mathcal{T}(s_{i,t}, a_{i,t}, s')\mathbb{E}[V_r(s')]. \\

@@

We begin by removing terms independent of $\mathbf{m}$:
\[
  \begin{split}
    \frac{\partial\mathcal{L}}{\partial\mathbf{m}} =
    &-\frac{1}{2}\dm\mathbb{E}[\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}] \\
    &+ \dm\left[\sum_{i=1}^N \sum_{t=1}^T \mathbb{E}[r(s_{i,t})] -
      \mathbb{E}[V_r(s_{i,t})] + \gamma\sum_{s' \in \mathcal{S}}
      \mathcal{T}(s_{i,t}, a_{i,t}, s')\mathbb{E}[V_r(s')]\right],
  \end{split}
\]
where
\begin{equation} \label{eq:1}
\begin{split} % TODO: dominated convergence theorem
    \dm\mathbb{E}[\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}] &=
    \dm\iiint\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}\mathcal{N}(\mathbf{u};
    \mathbf{m},
    \mathbf{S})q(\zeta)q(\bm\lambda)\,d\mathbf{u}\,d\zeta\,d\bm\lambda \\
    &= \iiint\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}\frac{\partial \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})}{\partial \mathbf{m}} q(\zeta)q(\bm\lambda)\,d\mathbf{u}\,d\zeta\,d\bm\lambda.
  \end{split}
\end{equation}
$\frac{\partial \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})}{\partial \mathbf{m}}$ can be calculated as
follows: % TODO: by the cookbook
\[
  \begin{split}
    \frac{\partial \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})}{\partial\mathbf{m}} &=
    \mathcal{N}(\mathbf{u}; \mathbf{m},
    \mathbf{S})\dm\left[-\frac{1}{2}(\mathbf{u} -
      \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} - \mathbf{m})\right] \\
    &= \mathcal{N}(\mathbf{u}; \mathbf{m},
    \mathbf{S})\left(-\frac{1}{2}\right)(\mathbf{S}^{-1} +
    \mathbf{S}^{-\intercal})(\mathbf{u} - \mathbf{m})\dm[\mathbf{u} -
    \mathbf{m}] \\
    &= \mathcal{N}(\mathbf{u}; \mathbf{m},
    \mathbf{S})\frac{1}{2}(\mathbf{S}^{-1} + \mathbf{S}^{-\intercal})(\mathbf{u}
    - \mathbf{m}).
  \end{split}
\]
Plugging it back to \eqref{eq:1} gives:
\[
  \begin{split}
    \dm\mathbb{E}[\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}] &=
    \iiint(\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u})\frac{1}{2}(\mathbf{S}^{-1} +
    \mathbf{S}^{-\intercal})(\mathbf{u} - \mathbf{m})\mathcal{N}(\mathbf{u};
    \mathbf{m},
    \mathbf{S})q(\zeta)q(\bm\lambda)\,d\mathbf{u}\,d\zeta\,d\bm\lambda \\
    &=
    \frac{1}{2}\mathbb{E}[(\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u})(\mathbf{S}^{-1}
    + \mathbf{S}^{-\intercal})(\mathbf{u} - \mathbf{m})]
  \end{split}
\]

@@

Similarly to the previous section,
\[
  \begin{split}
    \frac{\partial\mathcal{L}}{\partial\mathbf{S}} &=
    \frac{1}{2}\dS[\log|\mathbf{S}|]
    -\frac{1}{2}\dS\mathbb{E}[\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}] \\
    &+ \dS\left[\sum_{i=1}^N \sum_{t=1}^T \mathbb{E}[r(s_{i,t})] -
      \mathbb{E}[V_r(s_{i,t})] + \gamma\sum_{s' \in \mathcal{S}}
      \mathcal{T}(s_{i,t}, a_{i,t}, s')\mathbb{E}[V_r(s')]\right],
  \end{split}
\]
where % TODO: cite cookbook
\[ \dS[\log|\mathbf{S}|] =
  \frac{1}{|\mathbf{S}|}\frac{\partial|\mathbf{S}|}{\partial \mathbf{S}} =
  \frac{1}{|\mathbf{S}|}|\mathbf{S}|\mathbf{S}^{-\intercal} =
  \mathbf{S}^{-\intercal} \]
and
\[
  \begin{split}
    \dS\mathbb{E}[\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}] &=
    \dS\iiint\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}\mathcal{N}(\mathbf{u};
    \mathbf{m},
    \mathbf{S})q(\zeta)q(\bm\lambda)\,d\mathbf{u}\,d\zeta\,d\bm\lambda \\
    &=
    \iiint\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}\frac{\partial \mathcal{N}(\mathbf{u};
      \mathbf{m}, \mathbf{S})}{\partial \mathbf{S}}q(\zeta)q(\bm\lambda)\,d\mathbf{u}\,d\zeta\,d\bm\lambda.
  \end{split}
\]
$\frac{\partial \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})}{\partial \mathbf{S}}$ can be simplified to
\[
  \begin{split}
    \frac{\partial \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})}{\partial \mathbf{S}} &=
    \dS\left[\frac{1}{(2\pi)^{m/2}|\mathbf{S}|^{1/2}}\exp \left( -\frac{1}{2}
        (\mathbf{u} - \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} -
        \mathbf{m})\right)\right] \\
    &= \dS\left[\frac{1}{(2\pi)^{m/2}|\mathbf{S}|^{1/2}}\right]\exp \left( -\frac{1}{2}
        (\mathbf{u} - \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} -
        \mathbf{m})\right) \\
      &+ \frac{1}{(2\pi)^{m/2}|\mathbf{S}|^{1/2}}\dS\left[\exp\left( -\frac{1}{2}
        (\mathbf{u} - \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} -
        \mathbf{m})\right)\right] \\
    &=
    \frac{1}{(2\pi)^{m/2}}\dS\left[\frac{1}{|\mathbf{S}|^{1/2}}\right]\exp
    \left( -\frac{1}{2} (\mathbf{u} -
      \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} - \mathbf{m})\right) \\
    &- \frac{1}{2}\mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})\dS[(\mathbf{u}
    - \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} - \mathbf{m})], \\
  \end{split}
\]
where
\[ \dS\left[\frac{1}{|\mathbf{S}|^{1/2}}\right] =
  -\frac{1}{2}|\mathbf{S}|^{-3/2}\frac{\partial |\mathbf{S}|}{\partial \mathbf{S}} =
  -\frac{1}{2}|\mathbf{S}|^{-3/2}|\mathbf{S}|\mathbf{S}^{-\intercal} =
  -\frac{1}{2|\mathbf{S}|^{1/2}}\mathbf{S}^{-\intercal} \]
and
\[ \dS[(\mathbf{u} - \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} -
  \mathbf{m})] = -\mathbf{S}^{-\intercal}(\mathbf{u} - \mathbf{m})(\mathbf{u} -
  \mathbf{m})^\intercal\mathbf{S}^{-\intercal} \]
according to \textit{The Matrix Cookbook} \cite{petersen2008matrix}. Plugging
the two expression back in gives
\[ \frac{\partial \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})}{\partial
    \mathbf{S}} = -\frac{1}{2}\mathbf{S}^{-\intercal}\mathcal{N}(\mathbf{u};
  \mathbf{m}, \mathbf{S}) + \frac{1}{2}\mathcal{N}(\mathbf{u}; \mathbf{m},
  \mathbf{S})\mathbf{S}^{-\intercal}(\mathbf{u} - \mathbf{m})(\mathbf{u} -
  \mathbf{m})^\intercal\mathbf{S}^{-\intercal}. \]
Thus
\[ \dS\mathbb{E}[\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}] =
  \frac{1}{2}\mathbb{E}[(\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u})(\mathbf{S}^{-\intercal}(\mathbf{u}
  - \mathbf{m})(\mathbf{u} - \mathbf{m})^\intercal\mathbf{S}^{-\intercal} -
  \mathbf{S}^{-\intercal})]. \]

@@

\[
  \begin{split}
    \frac{\partial\mathcal{L}}{\partial\alpha} = &-
    \frac{1}{2}\da\mathbb{E}[\Tr[\Kuu^{-2}]] -
    \frac{1}{2}\da\mathbb{E}[\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}] -
    \frac{1}{2}\da\mathbb{E}[\log|\Kuu|] \\
    &+ \da [\alpha + \log\Gamma(\alpha) + (1 -
      \alpha)\psi(\alpha)] \\
    &+ \da \left[ \sum_{i=1}^N \sum_{t=1}^T \mathbb{E}[r(s_{i,t})] -
      \mathbb{E}[V_r(s_{i,t})] + \gamma\sum_{s' \in \mathcal{S}}
      \mathcal{T}(s_{i,t}, a_{i,t}, s')\mathbb{E}[V_r(s')] \right].
  \end{split}
\]
The first three terms can all be treated in the same way, since all three
expectations contain expressions without $\alpha$, but $\alpha$ appears in
$q(\zeta)$. Thus we can work with an abstract function as follows:
\[
  \begin{split}
    \da\mathbb{E}[f(k_\Theta, \mathbf{u})] &= \da \iiint f(k_\Theta, \mathbf{u})
    \frac{\beta^\alpha}{\Gamma(\alpha)} \zeta^{\alpha - 1} e^{-\beta\zeta}
    q(\bm\lambda) q(\mathbf{u})\,d\zeta\,d\bm\lambda\,d\mathbf{u} \\
    &= \iiint f(k_\Theta, \mathbf{u})
    \da \left[ \frac{\beta^\alpha}{\Gamma(\alpha)} \zeta^{\alpha - 1} \right]
      e^{-\beta\zeta}
    q(\bm\lambda) q(\mathbf{u})\,d\zeta\,d\bm\lambda\,d\mathbf{u}.
  \end{split}
\]
Then
\[
  \begin{split}
    \da \left[ \frac{\beta^\alpha}{\Gamma(\alpha)} \zeta^{\alpha - 1} \right] &=
    \frac{\da[\beta^\alpha\zeta^{\alpha - 1}]\Gamma(\alpha) -
      \beta^\alpha\zeta^{\alpha - 1}\Gamma'(\alpha)}{(\Gamma(\alpha))^2} \\
    &= \frac{\beta^\alpha\zeta^{\alpha - 1}\da[\alpha\log\beta + (\alpha -
      1)\log\zeta]\Gamma(\alpha) - \beta^\alpha\zeta^{\alpha -
        1}\Gamma'(\alpha)}{(\Gamma(\alpha))^2} \\
    &= \frac{\beta^\alpha\zeta^{\alpha - 1}(\log\beta + \log\zeta)\Gamma(\alpha)
      - \beta^\alpha\zeta^{\alpha - 1}\Gamma'(\alpha)}{(\Gamma(\alpha))^2} \\
    &= \frac{\beta^\alpha\zeta^{\alpha - 1}}{\Gamma(\alpha)}\left(\log\beta +
    \log\zeta - \frac{\Gamma'(\alpha)}{\Gamma(\alpha)}\right),
  \end{split}
\]
which means that
\[
  \begin{split}
    \da\mathbb{E}[f(k_\Theta, \mathbf{u})] &= \iiint f(k_\Theta, \mathbf{u})
    \frac{\beta^\alpha}{\Gamma(\alpha)}\zeta^{\alpha - 1}e^{-\beta\zeta} \left(
      \log\beta + \log\zeta - \frac{\Gamma'(\alpha)}{\Gamma(\alpha)} \right)
    q(\bm\lambda) q(\mathbf{u})\,d\zeta\,d\bm\lambda\,d\mathbf{u} \\
    &= \iiint f(k_\Theta, \mathbf{u}) \left(\log\beta + \log\zeta -
      \frac{\Gamma'(\alpha)}{\Gamma(\alpha)} \right) q(\zeta) q(\bm\lambda)
    q(\mathbf{u})\,d\zeta\,d\bm\lambda\,d\mathbf{u} \\
    &= \mathbb{E} \left[ f(k_\Theta, \mathbf{u}) \left(\log\beta + \log\zeta -
        \frac{\Gamma'(\alpha)}{\Gamma(\alpha)} \right) \right] \\
    &= \left( \log\beta - \frac{\Gamma'(\alpha)}{\Gamma(\alpha)}
    \right)\mathbb{E}[f(k_\Theta, \mathbf{u})] + \mathbb{E}[f(k_\Theta,
    \mathbf{u})\log\zeta].
  \end{split}
\]
Next,
\[ \da [\alpha + \log\Gamma(\alpha) + (1 - \alpha)\psi(\alpha)] =
  1 + \psi(\alpha) - \psi(\alpha) + (1 - \alpha)\psi'(\alpha) = 1 + (1 -
  \alpha)\psi'(\alpha). \]

>>>>

  \mathcal{L}(\bm\nu) &= \frac{1}{2}\log|\bm\Sigma| + \frac{1}{2}\log|\mathbf{S}| + \mathbb{E}\left[-\frac{1}{2}\Tr[\Kuu^{-2}] - \sum_{i=1}^d \log(\lambda_i + 1)\right] \\

@@

\cite{DBLP:books/lib/Bishop07}. Finally, we can simplify
$\sum_{i=1}^N\sum_{t=1}^Tr(s_{i,t})$ by choosing to represent all visited states
in $\mathbf{r} = (r_1, \dots, r_k)^\intercal$, and defining a new vector
$\mathbf{t} = (t_1, \dots, t_k)^\intercal$, where $t_i$ is the number of times
the state associated with the reward $r_i$ has been visited across all
demonstrations. Then
\mathbb{E} \left[ \sum_{i=1}^N\sum_{t=1}^Tr(s_{i,t}) \right] &=
    \mathbb{E}[\mathbf{t}^\intercal\mathbf{r}] =
    \mathbf{t}^\intercal\mathbb{E}[\mathbf{r}] =
    \mathbf{t}^\intercal\mathbb{E}_{(\Theta, \mathbf{u}) \sim
  q(\Theta)q(\mathbf{u})}[\Kru^\intercal\Kuu^{-1}\mathbf{u}] \\
  &= \mathbf{t}^\intercal\mathbb{E}_{\Theta \sim q(\Theta)}[\Kru^\intercal\Kuu^{-1}\mathbf{m}] = \mathbf{t}^\intercal\mathbb{E}[\Kru^\intercal\Kuu^{-1}]\mathbf{m}.
This allows us to simplify $\mathcal{L}(\bm\nu)$ to the following:

@@

  &+ \mathbf{t}^\intercal\mathbb{E}[\Kru^\intercal\Kuu^{-1}]\mathbf{m} - \sum_{i=1}^N \sum_{t=1}^T \mathbb{E}[V_r(s_{i,t})] - \gamma\sum_{s' \in \mathcal{S}} \mathcal{T}(s_{i,t}, a_{i,t}, s')\mathbb{E}[V_r(s')]. \\

@@

We begin by removing terms independent of $\mathbf{m}$:
\[
  \begin{split}
    \frac{\partial\mathcal{L}}{\partial\mathbf{m}} =
    &-\frac{1}{2}\dm\mathbb{E}[\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}]
    + \dm[\mathbf{t}^\intercal\mathbb{E}[\Kru^\intercal\Kuu^{-1}]\mathbf{m}] \\
    &- \sum_{i=1}^N \sum_{t=1}^T \dm\mathbb{E}[V_r(s_{i,t})] -
      \gamma\sum_{s' \in \mathcal{S}} \mathcal{T}(s_{i,t}, a_{i,t},
      s')\dm\mathbb{E}[V_r(s')],
  \end{split}
\]
where
\begin{equation} \label{eq:1}
\begin{split} % TODO: dominated convergence theorem
    \dm\mathbb{E}[\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}] &=
    \dm\iiint\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}\mathcal{N}(\mathbf{u};
    \mathbf{m},
    \mathbf{S})q(\zeta)q(\bm\lambda)\,d\mathbf{u}\,d\zeta\,d\bm\lambda \\
    &= \iiint\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}\frac{\partial \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})}{\partial \mathbf{m}} q(\zeta)q(\bm\lambda)\,d\mathbf{u}\,d\zeta\,d\bm\lambda.
  \end{split}
\end{equation}
$\frac{\partial \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})}{\partial \mathbf{m}}$ can be calculated as
follows: % TODO: by the cookbook
\[
  \begin{split}
    \frac{\partial \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})}{\partial\mathbf{m}} &=
    \mathcal{N}(\mathbf{u}; \mathbf{m},
    \mathbf{S})\dm\left[-\frac{1}{2}(\mathbf{u} -
      \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} - \mathbf{m})\right] \\
    &= \mathcal{N}(\mathbf{u}; \mathbf{m},
    \mathbf{S})\left(-\frac{1}{2}\right)(\mathbf{S}^{-1} +
    \mathbf{S}^{-\intercal})(\mathbf{u} - \mathbf{m})\dm[\mathbf{u} -
    \mathbf{m}] \\
    &= \mathcal{N}(\mathbf{u}; \mathbf{m},
    \mathbf{S})\frac{1}{2}(\mathbf{S}^{-1} + \mathbf{S}^{-\intercal})(\mathbf{u}
    - \mathbf{m}).
  \end{split}
\]
Plugging it back to \eqref{eq:1} gives:
\[
  \begin{split}
    \dm\mathbb{E}[\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}] &=
    \iiint(\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u})\frac{1}{2}(\mathbf{S}^{-1} +
    \mathbf{S}^{-\intercal})(\mathbf{u} - \mathbf{m})\mathcal{N}(\mathbf{u};
    \mathbf{m},
    \mathbf{S})q(\zeta)q(\bm\lambda)\,d\mathbf{u}\,d\zeta\,d\bm\lambda \\
    &=
    \frac{1}{2}\mathbb{E}[(\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u})(\mathbf{S}^{-1}
    + \mathbf{S}^{-\intercal})(\mathbf{u} - \mathbf{m})]
  \end{split}
\]

@@

Similarly to the previous section,
\[
  \begin{split}
    \frac{\partial\mathcal{L}}{\partial\mathbf{S}} &=
    \frac{1}{2}\dS[\log|\mathbf{S}|]
    -\frac{1}{2}\dS\mathbb{E}[\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}] \\
    &- \dS \left[\sum_{i=1}^N \sum_{t=1}^T \mathbb{E}[V_r(s_{i,t})] -
      \gamma\sum_{s' \in \mathcal{S}} \mathcal{T}(s_{i,t}, a_{i,t},
      s')\mathbb{E}[V_r(s')] \right],
  \end{split}
\]
where % TODO: cite cookbook
\[ \dS[\log|\mathbf{S}|] =
  \frac{1}{|\mathbf{S}|}\frac{\partial|\mathbf{S}|}{\partial \mathbf{S}} =
  \frac{1}{|\mathbf{S}|}|\mathbf{S}|\mathbf{S}^{-\intercal} =
  \mathbf{S}^{-\intercal} \]
and
\[
  \begin{split}
    \dS\mathbb{E}[\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}] &=
    \dS\iiint\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}\mathcal{N}(\mathbf{u};
    \mathbf{m},
    \mathbf{S})q(\zeta)q(\bm\lambda)\,d\mathbf{u}\,d\zeta\,d\bm\lambda \\
    &=
    \iiint\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}\frac{\partial \mathcal{N}(\mathbf{u};
      \mathbf{m}, \mathbf{S})}{\partial \mathbf{S}}q(\zeta)q(\bm\lambda)\,d\mathbf{u}\,d\zeta\,d\bm\lambda.
  \end{split}
\]
$\frac{\partial \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})}{\partial \mathbf{S}}$ can be simplified to
\[
  \begin{split}
    \frac{\partial \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})}{\partial \mathbf{S}} &=
    \dS\left[\frac{1}{(2\pi)^{m/2}|\mathbf{S}|^{1/2}}\exp \left( -\frac{1}{2}
        (\mathbf{u} - \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} -
        \mathbf{m})\right)\right] \\
    &= \dS\left[\frac{1}{(2\pi)^{m/2}|\mathbf{S}|^{1/2}}\right]\exp \left( -\frac{1}{2}
        (\mathbf{u} - \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} -
        \mathbf{m})\right) \\
      &+ \frac{1}{(2\pi)^{m/2}|\mathbf{S}|^{1/2}}\dS\left[\exp\left( -\frac{1}{2}
        (\mathbf{u} - \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} -
        \mathbf{m})\right)\right] \\
    &=
    \frac{1}{(2\pi)^{m/2}}\dS\left[\frac{1}{|\mathbf{S}|^{1/2}}\right]\exp
    \left( -\frac{1}{2} (\mathbf{u} -
      \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} - \mathbf{m})\right) \\
    &- \frac{1}{2}\mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})\dS[(\mathbf{u}
    - \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} - \mathbf{m})], \\
  \end{split}
\]
where
\[ \dS\left[\frac{1}{|\mathbf{S}|^{1/2}}\right] =
  -\frac{1}{2}|\mathbf{S}|^{-3/2}\frac{\partial |\mathbf{S}|}{\partial \mathbf{S}} =
  -\frac{1}{2}|\mathbf{S}|^{-3/2}|\mathbf{S}|\mathbf{S}^{-\intercal} =
  -\frac{1}{2|\mathbf{S}|^{1/2}}\mathbf{S}^{-\intercal} \]
and
\[ \dS[(\mathbf{u} - \mathbf{m})^\intercal\mathbf{S}^{-1}(\mathbf{u} -
  \mathbf{m})] = -\mathbf{S}^{-\intercal}(\mathbf{u} - \mathbf{m})(\mathbf{u} -
  \mathbf{m})^\intercal\mathbf{S}^{-\intercal} \]
according to \textit{The Matrix Cookbook} \cite{petersen2008matrix}. Plugging
the two expression back in gives
\[ \frac{\partial \mathcal{N}(\mathbf{u}; \mathbf{m}, \mathbf{S})}{\partial
    \mathbf{S}} = -\frac{1}{2}\mathbf{S}^{-\intercal}\mathcal{N}(\mathbf{u};
  \mathbf{m}, \mathbf{S}) + \frac{1}{2}\mathcal{N}(\mathbf{u}; \mathbf{m},
  \mathbf{S})\mathbf{S}^{-\intercal}(\mathbf{u} - \mathbf{m})(\mathbf{u} -
  \mathbf{m})^\intercal\mathbf{S}^{-\intercal}. \]
Thus
\[ \dS\mathbb{E}[\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}] =
  \frac{1}{2}\mathbb{E}[(\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u})(\mathbf{S}^{-\intercal}(\mathbf{u}
  - \mathbf{m})(\mathbf{u} - \mathbf{m})^\intercal\mathbf{S}^{-\intercal} -
  \mathbf{S}^{-\intercal})]. \]

@@

\[
  \begin{split}
    \frac{\partial\mathcal{L}}{\partial\alpha} = &-
    \frac{1}{2}\da\mathbb{E}[\Tr[\Kuu^{-2}]] -
    \frac{1}{2}\da\mathbb{E}[\mathbf{u}^\intercal\Kuu^{-1}\mathbf{u}] -
    \frac{1}{2}\da\mathbb{E}[\log|\Kuu|] \\
    &+ \mathbf{t}^\intercal\da\mathbb{E}[\Kru^\intercal\Kuu^{-1}]\mathbf{m} +
    \da [\alpha + \log\Gamma(\alpha) + (1 - \alpha)\psi(\alpha)] \\
      &- \da \left[\sum_{i=1}^N \sum_{t=1}^T \mathbb{E}[V_r(s_{i,t})] -
        \gamma\sum_{s' \in \mathcal{S}} \mathcal{T}(s_{i,t}, a_{i,t},
        s')\mathbb{E}[V_r(s')] \right].
  \end{split}
\]
The first four terms can all be treated in the same way, since all four
expectations contain expressions without $\alpha$, but $\alpha$ appears in
$q(\zeta)$. Thus we can work with an abstract function as follows:
\[
  \begin{split}
    \da\mathbb{E}[f(k_\Theta, \mathbf{u})] &= \da \iiint f(k_\Theta, \mathbf{u})
    \frac{\beta^\alpha}{\Gamma(\alpha)} \zeta^{\alpha - 1} e^{-\beta\zeta}
    q(\bm\lambda) q(\mathbf{u})\,d\zeta\,d\bm\lambda\,d\mathbf{u} \\
    &= \iiint f(k_\Theta, \mathbf{u})
    \da \left[ \frac{\beta^\alpha}{\Gamma(\alpha)} \zeta^{\alpha - 1} \right]
      e^{-\beta\zeta}
    q(\bm\lambda) q(\mathbf{u})\,d\zeta\,d\bm\lambda\,d\mathbf{u}.
  \end{split}
\]
Then
\[
  \begin{split}
    \da \left[ \frac{\beta^\alpha}{\Gamma(\alpha)} \zeta^{\alpha - 1} \right] &=
    \frac{\da[\beta^\alpha\zeta^{\alpha - 1}]\Gamma(\alpha) -
      \beta^\alpha\zeta^{\alpha - 1}\Gamma'(\alpha)}{(\Gamma(\alpha))^2} \\
    &= \frac{\beta^\alpha\zeta^{\alpha - 1}\da[\alpha\log\beta + (\alpha -
      1)\log\zeta]\Gamma(\alpha) - \beta^\alpha\zeta^{\alpha -
        1}\Gamma'(\alpha)}{(\Gamma(\alpha))^2} \\
    &= \frac{\beta^\alpha\zeta^{\alpha - 1}(\log\beta + \log\zeta)\Gamma(\alpha)
      - \beta^\alpha\zeta^{\alpha - 1}\Gamma'(\alpha)}{(\Gamma(\alpha))^2} \\
    &= \frac{\beta^\alpha\zeta^{\alpha - 1}}{\Gamma(\alpha)}\left(\log\beta +
    \log\zeta - \frac{\Gamma'(\alpha)}{\Gamma(\alpha)}\right),
  \end{split}
\]
which means that
\[
  \begin{split}
    \da\mathbb{E}[f(k_\Theta, \mathbf{u})] &= \iiint f(k_\Theta, \mathbf{u})
    \frac{\beta^\alpha}{\Gamma(\alpha)}\zeta^{\alpha - 1}e^{-\beta\zeta} \left(
      \log\beta + \log\zeta - \frac{\Gamma'(\alpha)}{\Gamma(\alpha)} \right)
    q(\bm\lambda) q(\mathbf{u})\,d\zeta\,d\bm\lambda\,d\mathbf{u} \\
    &= \iiint f(k_\Theta, \mathbf{u}) \left(\log\beta + \log\zeta -
      \frac{\Gamma'(\alpha)}{\Gamma(\alpha)} \right) q(\zeta) q(\bm\lambda)
    q(\mathbf{u})\,d\zeta\,d\bm\lambda\,d\mathbf{u} \\
    &= \mathbb{E} \left[ f(k_\Theta, \mathbf{u}) \left(\log\beta + \log\zeta -
        \frac{\Gamma'(\alpha)}{\Gamma(\alpha)} \right) \right] \\
    &= \left( \log\beta - \frac{\Gamma'(\alpha)}{\Gamma(\alpha)}
    \right)\mathbb{E}[f(k_\Theta, \mathbf{u})] + \mathbb{E}[f(k_\Theta,
    \mathbf{u})\log\zeta].
  \end{split}
\]
Next,
\[ \da [\alpha + \log\Gamma(\alpha) + (1 - \alpha)\psi(\alpha)] =
  1 + \psi(\alpha) - \psi(\alpha) + (1 - \alpha)\psi'(\alpha) = 1 + (1 -
  \alpha)\psi'(\alpha). \]
